\vspace{-0.5cm} XSGRB - The X-shooter GRB afterglow sample

### 2018-01-22 document/XSGRB_sample.tex
Comments by Jure

\abstract{In this work we present the spectra of all $\gamma$-ray bursts (GRBs)
	afterglows that have been observed at the VLT/X-shooter up to \termdate. This
	totals 137 spectroscopic observations of 118 individual bursts, of which
	redshifts have been secured for 91 \% covering a redshift range from 0.059 to
	$\sim$ 8.5. Some of these observation are late-time observations of the
	underlying host galaxy. Based on an set of selection criteria that are unbiased
	with regards to intrinsic properties of the GRBs, the follow-up effort has been
	guided by the \textit{Swift}-telescope to produce a homogeneous sample of
	afterglows for 86 spectra. We here provide a public release of all the reduced
	spectra, including continuum estimates and telluric corrections. We provide an
	assessment of the degree of completeness with respect to the parent GRB
	population, in terms of the X-ray properties of the bursts in the sample and
	find that the sample presented here is consistent with being drawn from the
	\textit{Swift} sample. For all bursts (39) for which it is possible, we provide
	a measurement of the hydrogen column density, increasing the total number of
	published HI column density measurements by $\sim$ 50 per cent. We constrain the
	fraction of dark bursts to be less than 28 per cent based on the detection of an
	optical afterglow and tentatively confirm previous results that larger optical
	darkness is correlated with increased X-ray absorption. This dataset provides a
	unique possibility to study the ISM across cosmic time. }

@@

In the beginning of the "{\it Swift} era" most of the follow-up spectroscopy was
secured using low-resolution spectrographs \citep[typically
R=$\lambda/\Delta\lambda$$<$1000, e.g.][]{Fynbo2009}. Low-resolution
spectroscopy is powerful for a number of reasons, first of all is the ability to
secure information for even very faint targets. This allows the measurements of
a number of important parameters such as redshifts, spectral slopes, and
extinction. For a handful of very bright afterglows high-resolution (typically
R$>$20000) spectra were secured, and for these events much more information
about conditions inside the host galaxies were extracted
\citep[e.g.,][]{Fiore2005, Thone2007, Prochaska2007, Vreeswijk2007,
	Castro-Tirado2010}.

The X-shooter spectrograph \citep{Vernet2011} is the first of the 2nd generation
instruments at the ESO Very Large Telescope (VLT). It is a high-throughput
echelle spectrograph covering the full spectral range from the UV atmospheric
cutoff around 300 nm to the near-infrared (near-IR) K-band ending around 2500 nm
with three spectral arms (UVB, VIS, NIR). It was designed very much with
transient source follow-up in mind as the fading luminosities of such sources
makes it urgent to secure as extensive coverage as possible in the shortest
possible time. At the same time, the resolution was designed to be in the range
4000--9000 in order to be able to get a large useful spectral range between the
many sky-background emission lines in the red and near-IR spectral ranges.

@@

Our ability to observe GRB afterglows is strongly dependent on the timing and
the precision of the target positions delivered by the triggering facilities. By
selecting only bursts that have been triggered on board the \textit{Swift} space
telescope \citep{Gehrels2004}, based on the Burst Alert Telescope (BAT)
\citep{Barthelmy2005}, we  start out with a homogeneously selected sample where
burst characteristics are delivered immediately, allowing for an informed
follow-up strategy. Despite the complexity of the triggering mechanism on board
\textit{Swift} \citep{Band2006, Coward2013a}, attempts at inferring properties
of the underlying GRB population based on the detection thresholds and
triggering algorithms have been made \citep{Lien2014, Graff2016}. Restricting
the follow-up effort to bursts detected by \textit{Swift}, we therefore ensure
that the limitations of the parent sample is well studied.

@@

To ensure a minimum of observability, we require that the GRB is visible from
the telescope site at Cerro Paranal, Chile, for a least 1 hour after the trigger
with the sun below $-12$ degrees. This secures time for the spectroscopic
observations to completed. Since the GRB population is isotropically distributed
on the sky as seen from Earth and because the GRB properties does not depend on
position on the sky \citep{Briggs1996}, this cut does not influence our ability
to fairly sample the underlying GRB population. The same arguments applies to
the requirement that there are no nearby bright stars. We additionally require
that the Galactic extinction is below $A_V \lesssim 0.5$ mag, based on the
extinction maps by \citet{Schlegel1998}. Because of isotropy, these additional
cuts should not influence the optical properties of the bursts themselves, only
our ability to successfully secure observations that allows us to investigate
the spectroscopic properties of GRBs.

We further discuss the effect of these selection criteria and their implication
for the compleness of the sample in Sect. \ref{completeness}.

@@

For 8 of the observed GRBs released here, RRM triggered. One of the RRM triggers
is outside the statistical sample and two of the RRM trigger are on short GRBs.
The fastest response between \swift~GRB trigger and the beginning of
spectroscopic integration is for GRB~160410A, for which the delay was only 8.4
minutes.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Observations} \label{obs}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure*}
	\centerline{\includegraphics{figures/skymap.pdf}} \caption{Mollweide projection
	in galactic coordinates of the full sky showing the positions on the sky of the
	bursts presented in this work. The equator is the Galactic plane. The empty
	circles are the positions of all the \textit{Swift} bursts detected until the
	end of 2016. In blue is shown the position of the bursts fulfilling the sample
	criteria specified in Sect. \ref{samplecrit}. The red circles are added around
	GRBs which have been followed up by X-shooter since the commissioning of the
	instrument. Blue points with red circles represent GRBs that enter our sample
	and have X-shooter spectroscopy. The different samples are compared in Sect.
	\ref{results}. The background shows the dust maps presented in
	\citet{Schlegel1998} where the sample criteria cut with $A_V \lesssim 0.5$ mag
	is removed. The colorbar below indicates the value of $A_V$ on the plot. The
	dotted lines indicate intervals of 30$^\circ$ in longitude and latitude.}
\label{fig:skymap}
\end{figure*}

@@

For a typical observation, each of the exposures in the nodding sequence has
been reduced as single observation and then subsequently combined to form a
single image. Because this strategy is employed, we can reject outliers in the
stack and weight by an averaged measure of the inverse variance of the
background. When weighting images where the noise in each pixel is dominated by
Poisson noise it important to estimate the background variance in a large enough
region, so that any correlation between the signal and the weights are removed.
To this end, the weight map is generated by a running median window over the
variance map, where the trace as been masked and width of the window is chosen
to be wide enough for median variance to be generated on the basis of several
hundred pixels. This weighting scheme automatically also optimally combines
images of different exposure times or images where the background is varying,
which is often the case when a burst has been observed close to twilight

@@

X-shooter, being installed at the VLT Cassegrain focus is  prone to flexures
during operations. The flexures modify the projection of the slit on the
detector with respect to the one obtained in daytime calibration. This requires
a modification of the wavelength solution in order to process correctly the
night-time data. Part of this correction is performed by the pipeline using the
frames taken during X-shooter Active Flexure Compensation procedure
\footnote{X-shooter User Manual available at
	https://www.eso.org/sci/facilities/paranal/instruments/xshooter/doc.html}. The
remaining offset is corrected by cross-correlating the observed sky spectrum
with a synthetic sky spectrum \citep{Noll2012, Jones2013} after the continuum,
estimated as the mode of all flux values, has been subtracted. To get the correct
seeing PSF with which to convolve the synthetic sky an initial refinement of the
wavelength solution have been obtained by cross-correlating the observed sky
with an unconvolved synthetic sky. This preliminary wavelength calibration is
applied to the observed sky. The synthetic spectrum is then convolved with an
increasing seeing PSF and the width that minimizes $\chi^2$ with the updated
observed sky is chosen to be the effective sky-PSF. Using the synthetic sky with
the matched resolution, a final wavelength calibration can then be calculated by
cross-correlating the observed sky with the correctly broadened sky spectrum, as
a function of a velocity offset. Both a multiplicative and an additive offset to
the wavelength calibration has been tested, but in terms of $\chi^2$, the model
with only a multiplicative offset is preferred.

The resulting offsets, which were smaller than 0.01 nm in the UVB and VIS data
and smaller than 0.05 nm in the NIR spectra, but changing over short period of
time were applied to the corresponding spectra.
Using the convolved synthetic sky, the pixels containing the brightest sky lines
have been added to the bad pixel map.

@@

The highest priority method is using the GRB afterglow continuum itself, where
the atmospheric conditions have directly been imprinted on the spectrum. The
telluric features can directly be fit with an atmospheric model
\citep{Smette2015, Kausch2015}, which can then be used to correct for the
absorption. The accuracy of the correction depends on the S/N for the target
spectrum, where we have chosen the requirement that the afterglow continuum
spectrum has a median signal-to-noise higher than a value of 10.

@@

In the last case, where the object is neither bright enough, or there for some
reason have not been observed a telluric standard, we rely on the synthetic sky
model by \citep{Noll2012, Jones2013} for which we generate a synthetic
transmission spectrum, which we then use, where the ambient parameters for the
observations have been used.

\begin{figure*}
	\centerline{\includegraphics[width=0.85\linewidth]{figures/GRB121024A.pdf}}
\caption{Telluric corrected, normalized spectrum of GRB~121024A at $z = 2.300$,
	that illustrates the typical data quality. The continuum estimate is shown in
	dashed red and the error spectrum in solid blue. The acquisition magnitude is
	$r$ = 20, meaning it is in the brighter end of the sample presented here, but
	not the brightest. The spectrum is rich in absorption, including absorption
	from molecular $\mathrm{H_2}$. The absorption trough visible at $\sim$ 4000
	$\AA$ is due to \lya~in the host. We have overplot the most prominent lines
	seen in GRB afterglows from \citet{Christensen2011a}. Additionally, three
	intervening systems are seen in this sightline This spectrum has been analyzed
	in detail in \citet{Friis2015}.} \label{fig:spectrum}
\end{figure*}

@@

We additionally provide an estimate of the continuum for all the spectra
presented here. For this, we have developed an algorithmic approach that
attempts to automatically finds the continuum placement along with the error on
the continuum estimate through an iterative procedure. The method is entirely
data-driven and does not rely on any physical assumptions. The method is applied on
each arm separately for each spectrum.

To estimate the continuum, a number of points (typically on the order of 100)
are inserted at random positions along the wavelength direction, the
flux-density of each point is determined by the median value of the spectrum in
a small region ($\sim 1 \mathrm{\AA}$) surrounding each point. The points are
fitted with a low order polynomial (we use \texttt{numpy.polynomial.chebyshev})
and iteratively, the point furthest away are removed until the polynomial fit
differ from the points by less than $\sim$ 5 per cent. This filtering is used
because the afterglow continuum of GRBs are power laws \citep{Piran2005}, and by
removing points that differ significantly from a smooth continuum shape will
guide the continuum estimate to a shape more reasonable for GRB afterglows.
Additionally, points spaced less than around 1 per cent of the entire spectral
coverage are pruned. The remaining filtered and pruned points are then spline
interpolated using \texttt{scipy.interpolate.splrep;splev}, which serves as a
first estimate of the continuum placement. An attempt to identify absorption and
emission regions are then carried out, where they are marked as such if the
difference between the estimated continuum and the observed spectrum is larger
than 3 - 5 times the associated error spectrum. All regions marked as affected
are then masked.

@@

The afterglow spectra described in this paper are obtained in
Target-of-Opportunity (override) mode. In most cases there is therefore little
possibility to tweak slit widths to the seeing at the time of observations (i.e.
to optimize spectral resolution and signal to noise), and almost all our data is
therefore taken with a fixed set of slit widths and binning, described above. In
a fair number of cases, the seeing full width at half maximum (FWHM) is
considerably smaller than the silt width, and the delivered spectral resolution
will then be determined by the seeing rather than slit width, as afterglows are
point sources (this is evidently not the case for extended sources, e.g for host
galaxies). The delivered resolution for slit width dominated spectra
post-reduction and extraction can easily be determined from the bright sky
emission lines. For afterglow spectra with very high signal to noise, the
delivered spectral resolution can at times be determined from the science data
themselves. However, in the presence of multiple velocity components in
absorption, other forms of line broadening, and a lack of lines at some
redshifts, this is difficult to do at poorer signal to noise ratios (the
majority of spectra in our sample). A broad starting value for the expect
resolution will help fitting of these spectra, and can be important in upper
limit determination, and for this reason we construct a crude relation between
the seeing and the delivered resolution at our slit width, binning, and
reduction pipeline settings.

To this end we use observations of telluric standard stars that are taken with
identical instrument settings as our afterglow spectra, usually just after the
science data, as part of the ESO X-shooter calibration plan. These spectra have
been reduced together with the afterglow spectra, using identical pipeline
settings with the same version of the pipeline. First we fit a Gaussian function
in the spatial direction of the trace of the standard star at 792 nm (i.e. in
the VIS arm). After this, we fit a series of  20 telluric absorption lines in
the telluric standard star spectra with Gaussians, taking care to select
transitions that are not almost-resolved multiples, should be intrinsically
unresolved, and are in areas with well defined continuum flux. We pick 34
telluric standard stars spanning a range of DIMM seeing values, with the
majority between $0.5-1.5 $\arcsec. The resulting distribution of spectral FWHM
(km/s) as a function of spatial FWHM at 792 nm is fairly well described by a
linear relation $a + b*x$, with $x$ the spatial FWHM in pixels (with 0.15 arc
sec per pixel),  $a= 21.4 \pm 1.3$ km/s, $b=1.4 \pm 0.2$. We use this linear
relation as a way to estimate the spectral resolution for medium to poor signal
to noise afterglow spectra in the VIS arm. To extend this to the UVB and NIR
arm, we measured a series of lines in NIR arm spectra of a subset of 19 sources
used for the VIS arm above, and find that the resulting distribution is
consistent with a simple scaling of the VIS arm relation by the ratio of
resolutions of the NIR and VIS arm for unresolved, slit filling, sources as
given on the ESO instrument website. The UVB arm contains no suitable absorption
lines to use, and we therefore use a scaled value as in the NIR arm. While this
simple method is not terribly accurate (for one, the spatial profile of the
trace is not a perfect Gaussian), but it gives a sufficiently accurate estimate
for the analysis of these poor signal to noise science spectra.

@@

All the spectra are made available at
\url{http://www.dark-cosmology.dk/~jselsing/XSGRB} and additionally through the
ESO archive in the form of phase 3 material
\footnote{\url{http://archive.eso.org/wdb/wdb/adp/phase3_main/form}}. This
release includes the spectra of GRBs outside the statistical sample and includes
both prompt afterglow observations as well as late time observations of the
associated hosts and represents \textit{all} afterglow spectra of GRBs carried
out by VLT/X-shooter from the commissioning of the instrument, \startdate, and
until the end of the last period of the program 098.A-0055, \termdate and thus
constitutes 8 years of GRB afterglow observations with VLT/X-shooter. An
overview of all the spectra provided here are given in Table.
\ref{tab:sample_overview}. For each burst, each individual observation is provided
in a separate reduction and in cases where observations have been repeated for
an increased S/N or to follow the temporal evolution, a combined spectrum is
also provided. No attempt has been made to join the spectroscopic arms, so for
each observation, three spectra are provided in separate files.

@@

Based on the sample criteria specified in Sect. \ref{samplecrit}, a total of 158
bursts have triggered \textit{Swift} in the period since the commissioning of
VLT/X-shooter. This sample constitutes the \textit{statistical sample} from
which we will derive statistical properties of the GRB population. The redshift
completeness of the full statistical sample is 60 per cent. We return to the
question of redshift completeness in in Sect. \ref{redshift}. From this sample,
90 GRBs have been spectroscopically followed up with X-shooter. In order to
assess whether the subset of bursts followed up are representative of the
underlying GRB parent population, we compare intrinsic properties of GRBs in our
sample to GRBs in the full sample followed up by \textit{Swift}. We show the
comparison between the BAT (15-150 keV) fluence, the XRT flux (0.3 - 10 keV) at
11 hours, and the intrinsic X-ray derived equivalent hydrogen column density,
$N_{\mathrm{HI,X}}$, in Fig. \ref{fig:swift_complete}. For the latter, we can
only use values of $N_{\mathrm{HI,X}}$ derived for bursts with a measured
redshift, excluding $\sim 75$ per cent of the bursts. We return to this last
point in Sect. \ref{darkness}.

@@

\begin{figure}
	\centerline{\includegraphics[width=9cm]{figures/XSGRB_p_values.pdf}}
	\caption{Relational graph shown the respective p-values that the different
	sample are drawn from the same underlying distribution. The arrows represent
	the comparison with each of the p-values are listed next to. Only in the HI
	column density distribution is there mild evidence against the null, but the
	discrepancy is mainly driven by a relatively larger fractional contribution
	from low-column hosts.}
	\label{fig:p_values}
\end{figure}

@@

Out of the 158 bursts in the statistical sample, 32($\sim$ 20 percent) have not
been observed due to reasons unrelated to the GRB or afterglow properties. The
reasons includes unavailability of the telescope due to technical maintenance
(e.g., mirror re-coating), a visitor rejecting the TOO trigger, or bad weather.
Because this cut is unrelated to the GRB properties, it will not change the
statistical properties of the full sample. Removing all these bursts,
drastically improves the redshift completeness from 60 per cent to 83 per cent.

@@

One of the objectives of our follow-up campaign is to measure the redshift
distribution for a well-defined, observationally unbiased and statistical useful
sample of GRBs. The imposed selection criteria (see Sect. \ref{samplecrit})
ensure that the GRBs entering our homogeneous sample fairly represent  the
underlying population. The redshift distribution of such a sample holds valuable
information about the occurrence of GRBs through cosmic time
\citep{Jakobsson2012, Perley2016a}.

>>>>

\abstract{In this work we present the spectra of all $\gamma$-ray bursts (GRBs)
	afterglows that have been promptly observed at the VLT/X-shooter up to
	\termdate. This totals spectroscopic observations of 102 individual bursts
	observed with 48 hours of the GRB trigger. Redshifts have been measured for 96
	\% of these, covering a redshift range from 0.059 to 7.84. Based on an set of
	selection criteria that are unbiased with regards to intrinsic properties of the
	GRBs, the follow-up effort has been guided by the \textit{Swift}-telescope to
	produce a homogeneous sample of afterglows for 86 spectra. We here provide a
	public release of all the reduced spectra, including continuum estimates and
	telluric corrections. For completeness, we also provide reductions for the
	late-time observations of the underlying host galaxies. We provide an assessment
	of the degree of completeness with respect to the parent GRB population, in
	terms of the X-ray properties of the bursts in the sample and find that the
	sample presented here is consistent with being drawn from the \textit{Swift}
	sample. For all bursts (39) for which it is possible, we provide a measurement
	of the hydrogen column density, increasing the total number of published HI
	column density measurements by $\sim$ 50 per cent. We constrain the fraction of
	dark bursts to be less than 28 per cent based on the detection of an optical
	afterglow and tentatively confirm previous results that larger optical darkness
	is correlated with increased X-ray absorption. This dataset provides a unique
	possibility to study the ISM across cosmic time. }

@@

In the beginning of the "{\it Swift} era" most of the follow-up afterglow
spectroscopy was secured using low-resolution spectrographs \citep[typically
R=$\lambda/\Delta\lambda$$<$1000, e.g.][]{Fynbo2009}. Low-resolution
spectroscopy is powerful as it allows us to secure information even for very faint
targets. This allows the measurements of a number of important parameters such
as redshifts, spectral slopes, and extinction. For a handful of very bright
afterglows high-resolution (typically R$>$20000) spectra were secured, and for
these events much more information about conditions inside the host galaxies
were extracted \citep[e.g.,][]{Fiore2005, Thone2007, Prochaska2007,
	Vreeswijk2007, Castro-Tirado2010}.

The X-shooter spectrograph \citep{Vernet2011} is the first of the 2nd generation
instruments at the ESO Very Large Telescope (VLT). It is a high-throughput
echelle spectrograph covering the full spectral range from the UV atmospheric
cutoff around 300 nm to the near-infrared (near-IR) K-band ending around 2500 nm
with three spectral arms (UVB, VIS, NIR). It was designed very much with
transient source follow-up in mind as the fading luminosities of such sources
makes it urgent to secure as extensive coverage as possible in the shortest
possible time. At the same time, the resolution was designed to be in the range
4000--9000 in order to be able to get a large useful spectral range between the
many sky-background emission lines in the red and near-IR spectral ranges. The
large spectral coverage allows for the spectroscopic observations of the
highest redshift GRBs.

@@

Our ability to observe GRB afterglows is strongly dependent on the timing and
the precision of the target positions delivered by the triggering facilities. By
selecting only bursts that have been triggered on board the \textit{Swift} space
telescope \citep{Gehrels2004}, based on the Burst Alert Telescope (BAT)
\citep{Barthelmy2005}, we  start out with a homogeneously selected sample where
burst characteristics are delivered immediately, allowing for an informed
follow-up strategy. Despite the complexity of the triggering mechanism on board
\textit{Swift} \citep{Band2006, Coward2013a}, attempts at inferring properties
of the underlying GRB population based on the detection thresholds and
triggering algorithms have been made \citep{Lien2014, Graff2016}. Restricting
the follow-up effort to bursts detected by \textit{Swift}, we therefore ensure
that the limitations of the parent sample are well studied.

@@

To ensure a minimum of observability, we require that the GRB is visible from
the telescope site at Cerro Paranal, Chile, for a least 1 hour after the trigger
with the sun below $-12$ degrees. This secures time for the spectroscopic
observations to completed. Since the GRB population is isotropically distributed
on the sky as seen from Earth and because the GRB properties do not depend on
position on the sky \citep{Briggs1996}, this cut does not influence our ability
to fairly sample the underlying GRB population. The same arguments apply to
the requirement that there are no nearby bright stars. We additionally require
that the Galactic extinction is below $A_V \lesssim 0.5$ mag, based on the
extinction maps by \citet{Schlegel1998}. Because of isotropy, these additional
cuts should not influence the optical properties of the bursts themselves, only
our ability to successfully secure observations that allows us to investigate
the spectroscopic properties of GRBs.

We further discuss the effect of these selection criteria and their implication
for the completeness of the sample in Sect. \ref{completeness}.

@@

There 8 GRBs in the sample that were being observed with the RRM mode. One of
the RRM triggers is outside the statistical sample and two of the RRM trigger
are on short GRBs. The fastest response between \swift~GRB trigger and the
beginning of spectroscopic integration is for GRB~160410A, for which the delay
was only 8.4 minutes.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Observations} \label{obs}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure*}
	\centerline{\includegraphics{figures/skymap.pdf}} \caption{Mollweide projection
	in galactic coordinates of the full sky showing the positions on the sky of the
	bursts presented in this work. The equator is the Galactic plane. Star symbols
	indicate bursts with measured redshifts and the circles indicate those without.
	The empty stars/circles are the positions of all the \textit{Swift} bursts
	detected until the end of 2016. A blue filling of the symbol indicates the
	position of bursts fulfilling the sample criteria specified in Sect.
	\ref{samplecrit}. The red circles are added around GRBs which have been
	followed up by X-shooter since the commissioning of the instrument. Blue points
	with red circles represent GRBs that enter our sample and have X-shooter
	spectroscopy. The different samples are compared in Sect. \ref{results}. The
	background shows the dust maps presented in \citet{Schlegel1998} where the
	sample criteria cut with $A_V \lesssim 0.5$ mag is removed. The colorbar below
	indicates the value of $A_V$ on the plot. The dotted lines indicate intervals
	of 30$^\circ$ in longitude and latitude.} \label{fig:skymap}
\end{figure*}

@@

For a typical observation, each of the exposures in the nodding sequence has
been reduced as a single observation and then subsequently combined to form a
single image. Because this strategy is employed, we can reject outliers in the
stack and weight by an averaged measure of the inverse variance of the
background. When weighting images where the noise in each pixel is dominated by
Poisson noise it is important to estimate the background variance in a large enough
region, so that any correlation between the signal and the weights are removed.
To this end, the weight map is generated by a running median window over the
variance map, where the trace has been masked and width of the window is chosen
to be wide enough for median variance to be generated on the basis of several
hundred pixels. This weighting scheme automatically also optimally combines
images of different exposure times or images where the background is varying,
which is often the case when a burst has been observed close to twilight

@@

X-shooter, being installed at the VLT Cassegrain focus is  prone to flexures
during operations. The flexures modify the projection of the slit on the
detector with respect to the one obtained in daytime calibration. This requires
a modification of the wavelength solution in order to process correctly the
night-time data. Part of this correction is performed by the pipeline using the
frames taken during X-shooter Active Flexure Compensation procedure
\footnote{X-shooter User Manual available at
	https://www.eso.org/sci/facilities/paranal/instruments/xshooter/doc.html}. The
remaining offset is corrected by cross-correlating the observed sky spectrum
with a synthetic sky spectrum \citep{Noll2012, Jones2013} after the continuum,
estimated as the mode of all flux values, has been subtracted. To get the
correct seeing PSF with which to convolve the synthetic sky an initial
refinement of the wavelength solution have been obtained by cross-correlating
the observed sky with an unconvolved synthetic sky. This preliminary wavelength
calibration is applied to the observed sky. The synthetic spectrum is then
convolved with an increasing seeing PSF and the width that minimizes $\chi^2$
with the updated observed sky is chosen to be the effective sky-PSF. Using the
synthetic sky with the matched resolution, a final wavelength calibration can
then be calculated by cross-correlating the observed sky with the correctly
broadened sky spectrum, as a function of a velocity offset. Both a
multiplicative and an additive offset to the wavelength calibration has been
tested, but in terms of $\chi^2$, the model with only a multiplicative offset is
preferred. The resulting offsets, which were smaller than 0.01 nm in the UVB and
VIS data and smaller than 0.05 nm in the NIR spectra, but changing over short
period of time were applied to the corresponding spectra. Using the convolved
synthetic sky, the pixels containing the brightest sky lines have been added to
the bad pixel map.

@@

The highest priority method is using the GRB afterglow continuum itself, where
the atmospheric conditions have directly been imprinted on the spectrum. The
telluric features can directly be fit with an atmospheric model
\citep{Smette2015, Kausch2015}, which can then be used to correct for the
absorption. The accuracy of the correction depends on the S/N per pixel for the
target spectrum, where we have chosen the requirement that the afterglow
continuum spectrum has a median signal-to-noise higher than a value of 10.

@@

In the last case, where the object is neither bright enough, or there for some
reason have not been observed a telluric standard, we rely on the synthetic sky
model by \citep{Noll2012, Jones2013} for which we generate a synthetic
transmission spectrum,  where the ambient parameters for the
observations have been used.

\begin{figure*}
	\centerline{\includegraphics[width=0.85\linewidth]{figures/GRB121024A.pdf}}
\caption{Telluric corrected, normalized spectrum of GRB~121024A at $z = 2.300$,
	that illustrates the typical data quality. The continuum estimate is shown in
	dashed red and the error spectrum in solid blue. The acquisition magnitude is
	$r$ = 20, meaning it is in the brighter end of the sample presented here, but
	not the brightest. The spectrum is rich in absorption and emission, including
	absorption from molecular $\mathrm{H_2}$. The absorption trough visible at
	$\sim$ 4000 $\AA$ is due to \lya~in the host. We have overplot the most
	prominent lines seen in GRB afterglows from \citet{Christensen2011a}.
	Additionally, three intervening systems are seen in this sightline This
	spectrum has been analyzed in detail in \citet{Friis2015}.}
\label{fig:spectrum}
\end{figure*}

@@

We additionally provide an estimate of the continuum for all the spectra
presented here. For this, we have developed an algorithmic approach that
attempts to automatically estimate the continuum placement along with the error on
the continuum estimate through an iterative procedure. The method is entirely
data-driven and does not rely on any physical assumptions. The method is applied on
each arm separately for each spectrum.

To estimate the continuum, a number of points (typically on the order of 100)
are inserted at random positions along the wavelength direction, the
flux-density of each point is determined by the median value of the spectrum in
a small region ($\sim 1 \mathrm{\AA}$) surrounding each point. The points are
fitted with a low order polynomial (we use \texttt{numpy.polynomial.chebyshev})
and iteratively, the point furthest away are removed until the polynomial fit
differ from the points by less than $\sim$ 5 per cent. This filtering is used
because the afterglow continuum of GRBs are power laws \citep{Piran2005}, and by
removing points that differ significantly from a smooth continuum shape will
guide the continuum estimate to a shape more reasonable for GRB afterglows.
Additionally, points spaced less than around 1 per cent of the entire spectral
coverage are pruned. The remaining filtered and pruned points are then spline
interpolated using \texttt{scipy.interpolate.splrep.splev}, which serves as a
first estimate of the continuum placement. An attempt to identify absorption and
emission regions are then carried out, where they are marked as such if the
difference between the estimated continuum and the observed spectrum is larger
than 3 - 5 times the associated error spectrum. All regions marked as affected
are then masked.

@@

The afterglow spectra described in this paper are obtained in
Target-of-Opportunity (override) mode. In most cases there is therefore little
possibility to tweak slit widths to the seeing at the time of observations (i.e.
to optimize spectral resolution and signal to noise), and almost all our data is
therefore taken with a fixed set of slit widths and binning, described above. In
a fair number of cases, the seeing full width at half maximum (FWHM) is
considerably smaller than the silt width, and the delivered spectral resolution
will then be determined by the seeing rather than slit width, as afterglows are
point sources (this is evidently not the case for extended sources, e.g for host
galaxies). The delivered resolution for slit-width-dominated spectra
post-reduction and extraction can easily be determined from the bright sky
emission lines. For afterglow spectra with very high signal to noise, the
delivered spectral resolution can at times be determined from the science data
themselves. However, in the presence of multiple velocity components in
absorption, other forms of line broadening, and a lack of lines at some
redshifts, this is difficult to do at poorer signal to noise ratios (the
majority of spectra in our sample). A broad starting value for the expect
resolution will help fitting of these spectra, and can be important in upper
limit determination, and for this reason we construct a crude relation between
the seeing and the delivered resolution at our slit width, binning, and
reduction pipeline settings.

To this end we use observations of telluric standard stars that are taken with
identical instrument settings as our afterglow spectra, usually just after the
science data, as part of the ESO X-shooter calibration plan. These spectra have
been reduced together with the afterglow spectra, using identical pipeline
settings with the same version of the pipeline. First we fit a Gaussian function
in the spatial direction of the trace of the standard star at 792 nm (i.e. in
the VIS arm). After this, we fit a series of  20 telluric absorption lines in
the telluric standard star spectra with Gaussians, taking care to select
transitions that are not almost-resolved multiples, should be intrinsically
unresolved, and are in areas with well defined continuum flux. We pick 34
telluric standard stars spanning a range of DIMM seeing values, with the
majority between $0.5-1.5 $\arcsec. The resulting distribution of spectral FWHM
(km/s) as a function of spatial FWHM at 792 nm is fairly well described by a
linear relation $a + b*x$, with $x$ the spatial FWHM in pixels (with 0.15 arc
sec per pixel),  $a= 21.4 \pm 1.3$ km/s, $b=1.4 \pm 0.2$. We use this linear
relation as a way to estimate the spectral resolution for medium to poor signal
to noise afterglow spectra in the VIS arm. To extend this to the UVB and NIR
arm, we measured a series of lines in NIR arm spectra of a subset of 19 sources
used for the VIS arm above, and find that the resulting distribution is
consistent with a simple scaling of the VIS arm relation by the ratio of
resolutions of the NIR and VIS arm for unresolved, slit filling, sources as
given on the ESO instrument website. The UVB arm contains no suitable absorption
lines to use, and we therefore use a scaled value as in the NIR arm. While this
simple method is not terribly accurate (for one, the spatial profile of the
trace is not a perfect Gaussian), it nevertheless gives a sufficiently accurate
estimate for the analysis of these poor signal to noise science spectra.

@@

All the spectra are made available at
\url{http://www.dark-cosmology.dk/~jselsing/XSGRB} and additionally through the
ESO archive in the form of phase 3 material
\footnote{\url{http://archive.eso.org/wdb/wdb/adp/phase3_main/form}}. This
release includes the spectra of GRBs outside the statistical sample and includes
both prompt afterglow observations as well as late time observations of the
associated hosts and represents \textit{all} afterglow spectra of GRBs carried
out by VLT/X-shooter from the commissioning of the instrument, \startdate, and
until the end of the last period of the program 098.A-0055, \termdate~and thus
constitutes 8 years of GRB afterglow observations with VLT/X-shooter. An
overview of all the spectra provided here are given in Table.
\ref{tab:sample_overview}. For each burst, each individual observation is provided
in a separate reduction and in cases where observations have been repeated for
an increased S/N or to follow the temporal evolution, a combined spectrum is
also provided. No attempt has been made to join the spectroscopic arms, so for
each observation, three spectra are provided in separate files.

@@

Based on the sample criteria specified in Sect. \ref{samplecrit}, a total of 158
bursts have triggered \textit{Swift} in the period since the commissioning of
VLT/X-shooter. This sample constitutes the \textit{statistical sample} from
which we will derive statistical properties of the GRB population. The redshift
completeness of the full statistical sample is 60 per cent. We return to the
question of redshift completeness in Sect. \ref{redshift}. From this sample,
90 GRBs have been spectroscopically followed up with X-shooter. In order to
assess whether the subset of bursts followed up are representative of the
underlying GRB parent population, we compare intrinsic properties of GRBs in our
sample to GRBs in the full sample followed up by \textit{Swift}. We show the
comparison between the BAT (15-150 keV) fluence, the XRT flux (0.3 - 10 keV) at
11 hours, and the intrinsic X-ray derived equivalent hydrogen column density,
$N_{\mathrm{HI,X}}$, in Fig. \ref{fig:swift_complete}. For the latter, we can
only use values of $N_{\mathrm{HI,X}}$ derived for bursts with a measured
redshift, excluding $\sim 75$ per cent of the bursts. We return to this last
point in Sect. \ref{darkness}.

@@

\begin{figure}
	\centerline{\includegraphics[width=9cm]{figures/XSGRB_p_values.pdf}}
	\caption{Relational graph showing the respective p-values that the different
	samples are drawn from the same underlying distribution. The arrows represent
	the comparison with each of the p-values are listed next to. Only in the HI
	column density distribution is there mild evidence against the null, but the
	discrepancy is mainly driven by a relatively larger fractional contribution
	from low-column hosts.}
	\label{fig:p_values}
\end{figure}

@@

Out of the 158 bursts in the statistical sample, 32($\sim$ 20 percent) have not
been observed due to reasons unrelated to the GRB or afterglow properties. The
reasons include unavailability of the telescope due to technical maintenance
(e.g., mirror re-coating), a visitor rejecting the TOO trigger, or bad weather.
Because this cut is unrelated to the GRB properties, it will not change the
statistical properties of the full sample. Removing all these bursts,
drastically improves the redshift completeness from 60 per cent to 83 per cent.

@@

One of the objectives of our follow-up campaign is to measure the redshift
distribution for a well-defined, observationally unbiased and statistically useful
sample of GRBs. The imposed selection criteria (see Sect. \ref{samplecrit})
ensure that the GRBs entering our homogeneous sample fairly represent  the
underlying population. The redshift distribution of such a sample holds valuable
information about the occurrence of GRBs through cosmic time
\citep{Jakobsson2012, Perley2016a}.
